---
title: "ps3_quarto"
github: "https://github.com/nchungigelman/gov51-ps3"
format: pdf
editor: visual
---


```{r}
install.packages("glmnet")

library(tidyverse)
library(knitr)
library(glmnet)

compas <- read_csv("data/raw/compas_clean.csv")
```
```{r}
compas <- read_csv("data/raw/compas_clean.csv")

dim(compas)

```

1.2: 
The compas dataset has 17 variables and 7214 unique observations. The unit of observation is an individual defendent or recidivism case from ProPublica's COMPAS dataset.

```{r}
glimpse(compas)
mean(compas$recidivism, na.rm = TRUE)
table(compas$recidivism)
prop.table(table(compas$recidivism))
```
1.3:
a) The overall recidivism rate in the data is 0.451. In plainspeak, approximately 45.1% of the people in the dataset were rearrestred within two years..
b) The outcome variable here is binary. There are two possible values for recidivism: 1 (the person was rearrested within two years) and 0 (the person wasn't rearrested within two years). In a linear model, the prediction represents the predicted probability that recidivism = 1. Because the outcome is binary, predicted values should be between 0 and 1, but the model might produce predictions outside of this range.

```{r}
summary_table <- data.frame(
  mean = sapply(compas, function(x) mean(x, na.rm = TRUE)),
  sd = sapply(compas, function(x) sd(x, na.rm = TRUE))
)

summary_table
```

1.4:
The variables priors_count (4.88) and age (11.89) have the largest raw standard deviations, but this is mostly due to the fact that both variables are measured in whole number units. Of the variables measuring proportions, recidivism (0.50), juv_other_count (0.50), juv_misd_count (0.49), juv_fel_count (0.47), and age_25_45 (0.50), felony, (0.48) and black (0.50) have notably larger standard deiations compared to the rest. The variables asian (0.07), native_american (0.05), and has_juv_felony (0.19) have lower standard deviations.

```{r}
library(dplyr)
library(ggplot2)

compas <- read.csv("data/raw/compas_clean.csv")

plot_data <- compas %>%
  transmute(
    priors_group = case_when(
      priors_count == 0 ~ "0",
      priors_count %in% 1:2 ~ "1-2",
      priors_count %in% 3:5 ~ "3-5",
      priors_count %in% 6:10 ~ "6-10",
      priors_count >= 11 ~ "11+"
    ),
    recidivism = recidivism
  ) %>%
  group_by(priors_group) %>%
  summarise(recid_rate = mean(recidivism, na.rm = TRUE), .groups = "drop") %>%
  mutate(priors_group = factor(priors_group, levels = c("0", "1-2", "3-5", "6-10", "11+")))

ggplot(plot_data, aes(x = priors_group, y = recid_rate)) +
  geom_col() +
  labs(
    x = "Priors count",
    y = "Recidivism rate",
    title = "Recidivism rate by priors count group"
  )
```

1.5:
Unsurprisingly, we find an essentially linear positive relationship between the number of prior offenses and the recidivism rate. This makes intuitive sense. Someone who has committed more offenses in the past is more likely to do so in the future and face legal consequences as a result (i.e. recidivism)

```{r}
compas <- read.csv("data/raw/compas_clean.csv")

set.seed(51)
train_idx <- sample(seq_len(nrow(compas)), size = floor(0.8 * nrow(compas)))
train <- compas[train_idx, ]
test <- compas[-train_idx, ]

nrow(train)
nrow(test)
```

2.1
The training data set comprises 5771 observations, while the test data set comprises 1443 observations.

```{r}
ols_main <- lm(recidivism ~ ., data = train)
summary(ols_main)
```

2.2
(a) At the 5% level, the predictors age, female, priors_count, felony, has_juv_misd, and age_25_45 are statistically significant.
(b) Holding all covariates constant, a one-unit increase in priors_count is associated with a 0.03 increase in the predicted probability of recidivism, or about 3 percentage points.

train_pred <- predict(ols_main, newdata = train)
rmse_train <- sqrt(mean((train$recidivism - train_pred)^2))

```{r}
train_pred <- predict(ols_main, newdata = train)
rmse_train <- sqrt(mean((train$recidivism - train_pred)^2))

test_pred <- predict(ols_main, newdata = test)
rmse_test <- sqrt(mean((test$recidivism - test_pred)^2))

rmse_train
rmse_test
rmse_test - rmse_train
```

2.3
We find a very small gap in RMSE between the training data (0.4606) and the test data (0.4622) of approximately 0.0016. Since this difference is so small, it suggests our model isn't overfitting the data and that using 16 predictors on a dataset of this size is fine.

```{r}
set.seed(51)
folds <- sample(rep(1:5, length.out = nrow(train)))

rmse_cv <- numeric(5)

for (k in 1:5) {
  train_k <- train[folds != k, ]
  test_k  <- train[folds == k, ]

  model_k <- lm(recidivism ~ ., data = train_k)
  pred_k  <- predict(model_k, newdata = test_k)

  rmse_cv[k] <- sqrt(mean((test_k$recidivism - pred_k)^2))
}

rmse_cv
mean(rmse_cv)
```